---
title: 'Project 4: Movie Recommendation'
author: "Asiri Silva, Matthew Endo, Lowell Tyner"
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    theme: readable
    toc: yes
    toc_float: yes
date: "12/10/2021"
---


```{r}
library(reshape2)
library(recommenderlab)
```

## Overview

Movie recommendation is an interesting application of data science where a user is shown a list of movies that might interest him. This is widely used in video streaming platforms such as Netflix, Hulu, and Amazon Prime, where the user is shown movies under "You may also like" to help the user select a movie quicker than browsing through the entire catalog. In this report, we discuss multiple approaches for recommending movies. 

- Under System I, we discuss recommending movies without user input on what movies the user likes. System 1 will purely depend on the user's preferred genre.
- Under System II, we discuss recommending movies incorporating user input on the movies the user prefers.

## Data

We will use the MovieLens 1M dataset for the analysis and the models. The dataset contains about 1 million anonymous ratings of approximately 3,900 movies made by 6,040 MovieLens users who joined MovieLens in 2000.

The data contains three datasets: Ratings, Movies, and User information

```{r}
#loading the datasets

myurl = "https://liangfgithub.github.io/MovieData/"

ratings = read.csv(paste0(myurl, 'ratings.dat?raw=true'), 
                   sep = ':',
                   colClasses = c('integer', 'NULL'), 
                   header = FALSE)
colnames(ratings) = c('UserID', 'MovieID', 'Rating', 'Timestamp')


movies = readLines(paste0(myurl, 'movies.dat?raw=true'))
movies = strsplit(movies, split = "::", fixed = TRUE, useBytes = TRUE)
movies = matrix(unlist(movies), ncol = 3, byrow = TRUE)
movies = data.frame(movies, stringsAsFactors = FALSE)
colnames(movies) = c('MovieID', 'Title', 'Genres')
movies$MovieID = as.integer(movies$MovieID)

# convert accented characters
movies$Title = iconv(movies$Title, "latin1", "UTF-8")
# extract year
movies$Year = as.numeric(unlist(
  lapply(movies$Title, function(x) substr(x, nchar(x)-4, nchar(x)-1))))


users = read.csv(paste0(myurl, 'users.dat?raw=true'),
                 sep = ':', header = FALSE)
users = users[, -c(2,4,6,8)] # skip columns
colnames(users) = c('UserID', 'Gender', 'Age', 'Occupation', 'Zip-code')
```

Sample of Ratings Data

Description of Variables

- UserID: An unique identifier for the user
- MovieID: An unique identifier for the movie
- Rating: Movie rating in a 1-5 scale
- Timestamp - Timestamp for when the user rated the movie

```{r}
head(ratings)
```

Sample of movies data

Description of the variables

- MovieID: An unique identifier for the movie
- Title: Movie Title
- Genres: The genres the movie belongs to. The movie genres will have values from the following
	* Action
	* Adventure
	* Animation
	* Children's
	* Comedy
	* Crime
	* Documentary
	* Drama
	* Fantasy
	* Film-Noir
	* Horror
	* Musical
	* Mystery
	* Romance
	* Sci-Fi
	* Thriller
	* War
	* Western
- Year: The year the movie was released.

```{r}
head(movies)
```

Sample of users data

Description of the variables:

- UserID: An unique identifier for the user
- Gender: User Gender
- Age: User Age. Age is chosen from the following values
	*  1:  "Under 18"
	* 18:  "18-24"
	* 25:  "25-34"
	* 35:  "35-44"
	* 45:  "45-49"
	* 50:  "50-55"
	* 56:  "56+"
- Occupation: User's Occupation. Occupation is chosen from the following variables.
	*  0:  "other" or not specified
	*  1:  "academic/educator"
	*  2:  "artist"
	*  3:  "clerical/admin"
	*  4:  "college/grad student"
	*  5:  "customer service"
	*  6:  "doctor/health care"
	*  7:  "executive/managerial"
	*  8:  "farmer"
	*  9:  "homemaker"
	* 10:  "K-12 student"
	* 11:  "lawyer"
	* 12:  "programmer"
	* 13:  "retired"
	* 14:  "sales/marketing"
	* 15:  "scientist"
	* 16:  "self-employed"
	* 17:  "technician/engineer"
	* 18:  "tradesman/craftsman"
	* 19:  "unemployed"
	* 20:  "writer"

- Zip-Code: User ZipCode

```{r}
head(users)
```


## System I

In System 1, we discuss potential approaches for recommending movies based on the user’s genre.  This approach is suitable for instances where the system cannot interact with the user to know what movies the user likes and only knows the preferred genre.

### Approach 1 - Most popular movies

Our first approach is to display the user with the most popular movies for a given genre. We define most popular as the movies with the most number of 4 and 5 ratings. Our thinking behind choosing 4,5 is that it will result in good movies, and also, taking the number of ratings, we will factor in the popularity of the movie as well. We will only show a movie if there are alteast 100, 4 and 5 ratings to ensure sufficient number of users have rated the movie


```{r}
#implmentation

# Data prep

#creating flags for each movie genre
movies_sub = movies
for(mov_ in c("Action","Adventure","Animation","Children's","Comedy","Crime","Documentary","Drama","Fantasy","Film-Noir","Horror","Musical",
"Mystery","Romance","Sci-Fi","Thriller","War","Western")){
  movies_sub[mov_] = FALSE
  movies_sub[grepl(mov_,movies_sub$Genres),mov_] = TRUE
}

# getting rating summary for 4,5s
ratings_sub = ratings[ratings$Rating >= 4,]
movie_45_rating_count = aggregate(list(ratings_sub$Rating), by = list(ratings_sub$MovieID), FUN = length)
names(movie_45_rating_count) = c("MovieID","rating_count")

#appending rating summary for movies
movie_n_rating = merge(movie_45_rating_count, movies_sub,  on = "MovieID")

#function to return top 5 movies
top_n_most_popular <- function(movie_rating_df, genre, n = 5){
  s = movie_rating_df[ (movie_rating_df[genre] == TRUE) & (movie_rating_df$rating_count > 100),] #alteast 100 ratings
  s$Genre = genre
  return(s[order(-s$rating_count),][1:n,c("Title","MovieID","Genre","rating_count")])
}

#write.csv(movie_n_rating,file = "/home/asiri/uiuc/psl/project4/movies_for_sys1.csv", row.names = FALSE)
saveRDS(movie_n_rating,file = "/home/asiri/uiuc/psl/project4/movies_for_sys1.rdata")
#readr
```

Application of System I, approach 1. 

We can call the function passing the summarizing movie rating dataset, and a desired genre,  and a value for how many top movies we want the algorithm to return. 

```{r}
#Top 5 most popular action movies

top_n_most_popular(movie_n_rating, "Action",5)
```

```{r}
#Top 5 most popular Crime movies
top_n_most_popular(movie_n_rating, "Crime",5)
```

It is clear that our approach return the movies with the highest number of reviews(in our case, highest number of 4,5 reviews)



### Approach 2 - Highest Rated Movies

Our approach 2 for System I considers showing movies with highest average rating for movies with atleast 100 reviews. 

```{r}
#implmentation

# Data prep

#creating flags for each movie genre
movies_sub = movies
for(mov_ in c("Action","Adventure","Animation","Children's","Comedy","Crime","Documentary","Drama","Fantasy","Film-Noir","Horror","Musical",
"Mystery","Romance","Sci-Fi","Thriller","War","Western")){
  movies_sub[mov_] = FALSE
  movies_sub[grepl(mov_,movies_sub$Genres),mov_] = TRUE
}

# getting rating summary, the number of ratings as well as the average rating
ratings_sub = ratings
movie_rating_count = aggregate(list(ratings_sub$Rating), by = list(ratings_sub$MovieID), FUN = length)
names(movie_rating_count) = c("MovieID","rating_count")

movie_rating_avg = aggregate(list(ratings_sub$Rating), by = list(ratings_sub$MovieID), FUN = mean)
names(movie_rating_avg) = c("MovieID","avg_rating")

movie_rating_summary = merge( movie_rating_avg,movie_rating_count, on = "MovieID")

#appending rating summary for movies
movie_n_rating = merge(movie_rating_summary, movies_sub,  on = "MovieID")

#function to return top 5 movies
top_n_highest_rated <- function(movie_rating_df, genre, n = 5){
  s = movie_rating_df[ (movie_rating_df[genre] == TRUE) & (movie_rating_df$rating_count > 100),] #alteast 100 ratings
  s$Genre = genre
  return(s[order(-s$avg_rating),][1:n,c("Title","Genre","avg_rating","rating_count")])
}
```



Application of System I, approach 2. 

We can call the function passing the summarizing movie rating dataset, and a desired genre,  and a value for how many highest rated movies we want the algorithm to return. 

```{r}
#Top 5 highest rated popular action movies
top_n_highest_rated(movie_n_rating, "Action",5)
```

```{r}
#Top 5 highest rated Crime movies
top_n_highest_rated(movie_n_rating, "Crime",5)
```

It is clear that our approach return the movies with the highest average rating for the given genre

For the Shiny application, we used the Approach 1 - the most popular movie approach.

## System II

Collaborative filtering is a way to answer the question - will a specific user like a specific item (movie, in this case)? Here we look at two different collaborative filtering methods - user based collaborative filtering (UBCF), and item based collaborative filtering (IBCF).

In our System II, the only information we have about the current user is the ratings for various movies that we ask the user to input.  UBCF and IBCF take different approaches using this user entered information to make a decision about what movies to recommend to the current user.

Here we consider two approaches

- Item-Based Collaborative Filtering(IBCF)
- User-Based Collaborative Filtering(UBCF)


### Item-Based Collaborative Filtering(IBCF)

In item-based collaborative filtering is a model-based approach which produces recommendations based on the relationship between items inferred from the rating matrix. The assumption behind this approach is that users will prefer items that are similar to other items they like.

The model-building step consists of calculating a similarity matrix containing all item-to-item similarities using a given similarity measure. Pearson correlation and Cosine similarity are popular when it comes to similarity measures

### User-Based Collaborative Filtering(UBCF)

Some description on UBCF
Some description on UBCF
Some description on UBCF
Some description on UBCF



** Our ShinyApp uses UBCF to recommend movies based on the user input **



The the following section we build two models, IBCF and UBCF to evaluate accuracy on 10 random train/test samples

We will use the rating dataset, follows the steps mentioned in the `recommenderlab` documentation

```{r}
#dataset setup

#creating a rating matrix
rating_matrix = as.matrix(dcast(ratings, UserID ~ MovieID, value.var = "Rating"))

#let's conver the matrix to the realRatingMatrix
real_rating_matrix =  as(rating_matrix, "realRatingMatrix")
```

It is important to remove the rating bias for rating matrices. We use `normalize()` to achieve this.
We will use Z-score normalization

```{r}
normalize = normalize(real_rating_matrix,method="Z-score")
```

In the following code segment, we setup IBCF and UBCF, and evaluate the prediction accuracy on 10 sample runs

< give the defition of each para >

```{r}
#defining an evaluation scheme with 80%
ibcf_rmse = rep(0,10)
ubcf_rmse = rep(0,10)

for(i in 1:10){
  e <- evaluationScheme(real_rating_matrix[1:200], #JUST RUNNING A SAMPLE FOR NOW
                        method="split", 
                        train=0.8, 
                        given=15, 
                        goodRating=5,
                        )
  
  #building a recommender
  
  #Building a IBCF Recommender
  r1_ibcf <- Recommender(getData(e, "train"),
                         method = "IBCF",
                        param=list(normalize="Z-score",
                                   method="cosine",
                                   k = 30 #number of nearest neighbors
                        )
                         )
  
  #Building a UBCF Recommender
  r2_ubcf <- Recommender(getData(e, "train"),
                       method = "UBCF",
                      param=list(normalize="Z-score",
                                 method="cosine",
                                 nn = 25, #number of nearest neighbors ??
                                 weighted = TRUE
                      )
                       )
  #Evaluating accuracy IBCF
  pred_ibcf <- predict(r1_ibcf, getData(e, "known"), type="ratings")
  ibcf_rmse[i] = calcPredictionAccuracy(pred_ibcf, getData(e, "unknown"))[["RMSE"]]
  
  #Evaluating accuracy UBCF
  pred_ibcf <- predict(r2_ubcf, getData(e, "known"), type="ratings")
  ubcf_rmse[i] = calcPredictionAccuracy(pred_ibcf, getData(e, "unknown"))[["RMSE"]]
  
}
```

```{r}
ibcf_rmse
```

```{r}
ubcf_rmse
```


Sources:

- https://www.cs.carleton.edu/cs_comps/0607/recommend/recommender/itembased.html
- https://www.r-bloggers.com/2015/11/item-based-collaborative-filtering-recommender-systems-in-r/
- https://cran.r-project.org/web/packages/recommenderlab/vignettes/recommenderlab.pdf
